The FAIRification of research in real-world evidence: A practical introduction to reproducible analytic workflows using Git and R

### 2023-11-22 manuscript/manuscript.qmd
As per R#2 comment: added more commands (fetch, merge, rebase) and moved this as first occurrence in git workflow section, highlighted "time machine" aspect, moved reproducible computation environments to supplement

Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved. For our purposes, a repository can be seen as an isolated project directory which includes all neccessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).

@@

## Repository structure {#sec-repository-structure}

At this step, we now can populate the repository and start working on study-relevant documents and analyses. To enhance clarity, it is recommended to use a logical structure and an informative nomenclature for sub-directories and file names (a minimal example is illustrated in @fig-structure). To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file. This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added afterwards by creating a text file with the name `README.md` in the root directory of the repository.

@@

@@

As introduced before, Git allows researchers to keep track of changes by making snapshots of the repository every time the state of a project is saved. In Git terms, this fundamental step is referred to as a `commit`. Generally, it is up to the user when to commit, which files to commit and how the commit should be documented in a `commit message`.

@@

## Tracking changes

@@

To that end, Zenodo[@europeancommission.directorategeneralforresearchandinnovation.2019;@zenodo], a free and open-source platform funded by the European Commission and developed and maintained by European Organization for Nuclear Research (CERN), allows users to share diverse research artifacts. The platform seamlessly interfaces with GitHub, enabling the creation of referenceable DOIs for existing code repositories. The steps required to obtain a DOI involve the linkage of a Zenodo and a GitHub account[@zenodoGitHub]. Once linked, a user will be able to select a specific release of a repository which automatically registers a new DOI and a corresponding badge which can be included and referenced in a GitHub README file.

## Keeping track of R software versions and paths

A major advantage of modern programming languages is that they provide a large pool of open-source packages. Packages are compilations of functions that automate frequently used procedures (e.g., querying new user cohorts[@tidyverse] or propensity score matching[@MatchIt]) which expedites the implementation of RWE studies, mitigates copy-paste errors and makes code much more friendly to read and review.

A further step to improve reproducibility when using the R programming language is to ensure that anyone, intending to reproduce analytic results from a `Git` repository, is able to work off the same R package versions and dependencies without having to perform manual installations or path adjustments. These requirements are nothing exclusive to R, but apply to any type of programming language.

### Reproducible R computing environments

One way to achieve this is through self-contained computing environments via Docker images which are isolated and standardized environments which include everything needed to run and deploy applications and code regardless of a user's local operating system.[@boettiger2017a; @nüst2020] Another more lightweight, yet effective way is by using the `renv` R package[@renv], which helps creating **re**producible **env**ironments and manages R package dependencies for a given project. Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>). To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step necessary is running the following command in R.

```{r}
renv::restore()
```

Appending the session info to the end of an R script can provide further context about the dependencies and versions of an analytic script.

```{r}
sessionInfo()
```

### Absolute versus relative paths

Another common issue when trying to reproduce analytic code is the use of absolute paths such as in the following example, where we try to call an `icd_code` file from the project's `protocol` sub-directory.

```{r}
icd_codes <- read.csv('C:/User/Projects/rwe_study/protocol/icd_codes.csv')
```

The use of absolute paths will make it impossible for any collaborator to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming. A better solution to this is the use of relative paths. A popular R package to implement this is the `here` package[@here] which enables file referencing in project-oriented workflows. The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory). Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any collaborator re-running this script.

```{r}
library(here)
icd_codes <- read.csv(here("protocol", "icd_codes.csv"))
```

>>>>

Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved and hence can be seen as a "*time machine*" that enables researchers and developers to compare code and repositories across different versions over time. For our purposes, a repository can be seen as an isolated project directory which includes all necessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).

@@

In case a users wants to join or contribute to an already existing project, `git clone` works the same way with the only difference that the repository already contains some files and is not empty.

## Repository structure, README file and licenses {#sec-repository-structure}

At this step, we now can populate the repository and start working on study-relevant files and analyses. To enhance clarity, it is recommended to use a logical structure and an informative nomenclature for sub-directories and file names (a minimal example is illustrated in @fig-structure).

To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file. This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added afterwards by creating a text file with the name `README.md` in the root directory of the repository. It is critically important to use the `README.md` file to provide essential instructions and information about the repository and the corresponding project. This should include a brief summary about the project's background and objectives, instructions on how to install software or manage computation environments and dependencies, details regarding codebooks, important files and the organization of the repository, contribution guidelines, contact information and the project's license to outline the terms under which others can use, modify, and distribute the project. Especially for open-source software projects choosing a suitable license is critical for which GitHub provides helpful guidance and information under <https://choosealicense.com/>.

@@

### Synchronize changes from a remote repository to a local repository (`pull`, `fetch`, `merge` and `rebase`)  {#pull-fetch-merge}

Especially when working collaboratively with multiple users on a remote repository, it is usually good practice to synchronize the status of one's local repository and the remote repository before making any changes since there could be potential modifications and updates already made by others. To avoid any conflicts between two different versions and always work with the most recent version of a file, the first step of the git workflow involves a `git pull` or `git fetch` command.

-   `git fetch` downloads any changes made to a remote repository into a user's local repository. However, the command isolates the potentially modified files from the state of the existing local files. This is useful since it enables users to safely review changes using `git log` (we will cover this in more detail in @sec-tracking-changes) before integrating those locally. To eventually allow the integration of these changes locally, users can do so using the `git merge` or `git rebase` command. The difference between these two commands is how the history of the versions between different snapshots of the repository is tracked, for which `git merge` is more common, simpler and less error-prone and `git rebase` is a more advanced feature (for the purpose of this manuscript we will mostly focus on `git merge`).

```{bash}
# fetch ("download") updated version of a remote repository
git fetch

# use merge to integrate these changes into local files using merge
git merge
```

-   `git pull` combines `git fetch` and `git merge` in one command and directly downloads [and]{.underline} integrates all changes from a remote repository locally.

```{bash}
# download and integrate content from remote to local
git pull
```

@@

Git allows researchers to keep track of changes by making snapshots of the repository every time the state of a project is saved. In Git terms, this fundamental step is referred to as a `commit` and it enables users to compare code versions across snapshots. Generally, it is up to the user when to commit, which files to commit and how the commit should be documented in a `commit message`. However, it is best practice to *commit early and often*. That is, smaller and more granular commits that reflect a single unit of work (e.g., a change of selected confounders to adjust for) make it easier to comprehend the changes and associated analytic results and revert potential errors.

@@

## Tracking changes {#tracking-changes}

@@

To that end, Zenodo[@europeancommission.directorategeneralforresearchandinnovation.2019; @zenodo], a free and open-source platform funded by the European Commission and developed and maintained by European Organization for Nuclear Research (CERN), allows users to share diverse research artifacts. The platform seamlessly interfaces with GitHub, enabling the creation of referenceable DOIs for existing code repositories. The steps required to obtain a DOI involve the linkage of a Zenodo and a GitHub account[@zenodoGitHub]. Once linked, a user will be able to select a specific release of a repository which automatically registers a new DOI and a corresponding badge which can be included and referenced in a GitHub README file.
